Gentoo Package Dependencies over Time

### 2014-08-08 Paper/paper.tex
More work on section 5

The interpretation of the variables and parameters and their dimensions is presented in table~\ref{tbl:bass-model}. When applying the formula one should note that it is non-linear, so ordinary linear regression can not be used. Instead one can use non-linear least squares regression, but note the correct number of the degrees of freedom. This can be done using existing mathematical/statistical packages. For this thesis the  \verb|NonlinearModelFit| procedure in Mathematica was used.

@@

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to \eqref{eq:bass} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjust coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

The plots were drawn using a thick red line for the model and shades of red for the prediction bands. The thick red line was drawn according to \eqref{eq:bass}, with the mean values used as parameters. Then single value predictions bands were calculated for $90\%$, $95\%$, $99\%$ and $99.9\%$ confidence and were drawn in progressively darker shades of pink. They represent a prediction for where a single additional value would likely fall. According to the model and the uncertainty introduced by the fit, there is a $90\%$ chance that it will fall in the innermost band, $95\%$ chance that it would falil in the second band, etc. If the data fits the model properly, one would expect to see $90\%$ of the points in the inner band. Finally, the empirical data points were plotted as black dots, connected with a thin vertical line - the residual error of the model.

@@

\begin{center}
\begin{BVerbatim}
cvs -d :pserver:anonymous@anoncvs.gentoo.org\
  /var/cvsroot co -D 12/01/2003 gentoo-x86
\end{BVerbatim}
\end{center}

@@

@@

\begin{figure}[t]
\small\centering
\begin{BVerbatim}
>=media-libs/taglib-1.6.1[asf,mp4]
>=media-libs/taglib-extras-1.0.1
player? (
    app-crypt/qca:2
    >=app-misc/strigi-0.5.7[dbus,qt4]
    || ( >=dev-db/mysql-5.0.76
         =virtual/mysql-5.1 )
    >=kde-base/kdelibs-4.3[opengl?,semantic-desktop?]
    sys-libs/zlib
    x11-libs/qt-script
    >=x11-libs/qtscriptgenerator-0.1.0
\end{BVerbatim}
\caption{Fragment of the runtime dependencies of the Amarok music player.}\label{fig:amarokruntime}
\end{figure}

@@

The dataset collected is available at
\begin{center}
\href{http://datahub.io/dataset/gentoo-dependency-graph}{\texttt{http://datahub.io/dataset/gentoo-dependency-graph}}
\end{center}
as compressed GML graphs for every snapshot. Raw data is available at
\begin{center}
\url{http://www.utwente.nl/mb/iebis/staff/amrit/timeline.squashfs}\\
\end{center}
as a SquashFS compressed filesystem. The SquashFS format allows one to mount the compressed data as a read-only disk drive and operate directly on the dataset. It contains the pre-processed ebuild files and the file \verb|graphs.bip| that contains all of the graphs in a custom format specialized for quick processing of large graphs over time. A C++ toolkit to work with the \verb|graphs.bip| file is available at \url{https://github.com/Recmo/depgraph}.

In a separate paper \citep{rem142} we analyzed the changes in the dependency graph over time. In particular the growth of the number of dependers on a given package is explained using the Bass model of innovation diffusion.

@@

From the entire list of packages, a few well-known (at least according to the author) packages where selected. The selection criteria was that the package must not have existed before (around) 2004, because the Gentoo Portage database was still too immature then, and the package must have gained a considerable number of dependers since its introduction. In \ref{tbl:results}, we list some selected packages along with their parameters. In all cases the adjusted coefficient of determination $\overline{R}^2$ was more than $99\%$.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in \ref{fig:git}, the corresponding statistics are in \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ user to adopt \verb|git| out of sheer innovation. Taking the analogy with persons, if someone from the $450$ current non-users where to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Very much imitator driven!

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process, in particular the rapid steep growth and the stable user base afterwards. The name is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

\subsection{Hypothesis testing}

· Filter on >2004  >30
· Select 10 randomly where
	· External confirmation that the project was actually started at that point in time and not as a spin-off or refactor.
	· No 'noise' Δ >

Hypotheses: “Package dependency growth fits a Bass diffusion process”

To test this hypotheses we do the following:

→ Explain why discontinuities are unnatural and filter on them

\begin{itemize}
	\item Start with all the packages in the Gentoo portage database. (21\,380)
	\item Filter out the packages that existed before 2004-01-01. (14\,706)
	\item Filter out packages that do not have at least 20 dependees at some point. (324)
	\item Filter out packages labeled “-virtual”, “-virtuals”, “-proto” or “-meta”. (259)
	\item Randomly pick packages and manually verify that:
	\begin{enumerate}
		\item[A.] The data does not contain strong discontinuities.
		\item[B.] The package corresponds to an open source project.
		\item[C.] The package has not been forked of or factored out of another package.
		\item[D.] The first included version should be close to a '1.0' release.
		\item[E.] That open source project has not been in Gentoo as (part of) some other package.
	\end{enumerate}
	This is verified by consulting
	\begin{itemize}
		\item The changelog of the ebuild
		\item Any readme or changelog included in the source code (earliest available version).
		\item The projects revision history.
		\item The projects homepage.
	\end{itemize}
\end{itemize}

@@

No packages where rejected on the 'E' ground, but many of these are covered by the 'C' reason. No extensive check was done on co-occurence of reasons.

More information about a packages can be found at
\begin{center}
\href{https://packages.gentoo.org/package/dev-vcs/git}{\texttt{https://packages.gentoo.org/package/\textcolor{MidnightBlue}{dev-vcs/git}}}
\end{center}
where \texttt{\textcolor{MidnightBlue}{dev-vcs/git}} is replaced with the name of the package.

The growth curves with large discontinuities are rejected. The hypotheses concerns open source projects depending on other open source projects. This is reflected in the Gentoo Portage dataset as dependency relations between packages, but this reflection can be quite noisy. It regularly happens that Gentoo developers re-organize the package database and within a datapoint (i.e. a single month) add or drop a large number of dependees to a package. The changes can be as large as a complete drop to zero or suddenly tripling the number of dependees. If this is translated back to dependee behaviour, it would mean that in a short span of time a large number of projects decide to use or drop a certain technology. Since the dependee projects are assumed to operate (mostly) independently, this is very unlikely. Even in instances of deprecation, where a project gets discontinued and tells its dependees to move on, the resulting decline is far from instantaneous. Therefore, growth curves with large discontinuities are rejected.

@@

\begin{table}
\small\centering
\caption{Results of fitting the Bass model.}\label{tbl:results}
\begin{tabular}{lr@{ $\pm$}rr@{ $\pm$}rr@{ $\pm$}rr}
\toprule
Package & \multicolumn{2}{c}{$p$} & \multicolumn{2}{c}{$q$} & \multicolumn{2}{c}{$M$} & $\overline{R}^2$\\
\midrule
% 34, 65, 218, 77, 157, 240, 242, 17, 206, 143
\texttt{dev-dotnet/gconf-sharp}   & 0.168 & 0.212 & 0.754 & 0.434 &  21 &   1 & 99.61\%\\
\texttt{dev-haskell/cabal}        & 0.055 & 0.074 & 0.154 & 0.380 & 246 & 438 & 98.55\%\\
\texttt{dev-libs/libunique}       & 0.047 & 0.099 & 0.136 & 0.402 & 287 & 732 & 98.58\%\\
\texttt{dev-python/lxml}          & 0.007 & 0.009 & 0.488 & 0.217 &  53 &  53 & 98.82\%\\
\texttt{dev-python/simplejson}    & 0.011 & 0.029 & 0.463 & 0.459 &  97 & 148 & 98.98\%\\
\texttt{dev-vcs/git}              & 0.000 & 0.006 & 0.841 & 0.144 & 734 & 369 & 99.24\%\\
\texttt{dev-vcs/mercurial}        & 0.001 & 0.005 & 0.517 & 0.259 & 174 & 481 & 98.24\%\\
\texttt{gnome-base/gnome-keyring} & 0.026 & 0.036 & 0.833 & 0.133 &  42 &   2 & 99.66\%\\
\texttt{media-libs/libcanberra}   & 0.055 & 0.117 & 1.436 & 0.439 &  37 &   3 & 99.68\%\\
\texttt{x11-misc/xdg-utils}       & 0.050 & 0.043 & 0.257 & 0.242 & 115 & 127 & 99.56\%\\
\bottomrule
\end{tabular}
\end{table}

% TODO: Write something about discretization noise

% TODO:

The package \texttt{dev-libs/libunique} does not have a good fit, but looking at the graph, \raisebox{-.15em}{\includegraphics[height=.8em,trim=7em 0 0 0,clip=true]{sparkline-218-dev-libs-libunique.pdf}}, it goes down at the end. This is demonstraded even more clearly in \texttt{net-libs/xulrunner}, \raisebox{-.15em}{\includegraphics[height=.8em,trim=3em 0 0 0,clip=true]{sparkline-140-net-libs-xulrunner.pdf}}, which declines almost completely.

>>>>

The interpretation of the variables and parameters and their dimensions is presented in table~\ref{tbl:bass-model}. When applying the formula one should note that it is non-linear, so ordinary linear regression can not be used. Instead one can use non-linear least squares regression, but note the correct number of the degrees of freedom. This can be done using existing mathematical/statistical packages. For this paper the \verb|NonlinearModelFit| procedure in Mathematica was used.

@@

More information about a packages can be found at
\begin{align}
\mbox{
\href{https://packages.gentoo.org/package/dev-vcs/git}{\texttt{https://packages.gentoo.org/package/\textcolor{MidnightBlue}{dev-vcs/git}}}
}\label{url:pkg}
\end{align}
where \texttt{\textcolor{MidnightBlue}{dev-vcs/git}} is replaced with the name of the package.

@@

\begin{align}
\mbox{\texttt{cvs -d :pserver:anonymous@anoncvs.gentoo.org}}&\;\backslash \notag\\
\mbox{\texttt{  /var/cvsroot co -D 12/01/2003 gentoo-x86}}
\label{cmd:cvs}
\end{align}

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\small\centering
\begin{BVerbatim}
>=media-libs/taglib-1.6.1[asf,mp4]
>=media-libs/taglib-extras-1.0.1
player? (
    app-crypt/qca:2
    >=app-misc/strigi-0.5.7[dbus,qt4]
    || ( >=dev-db/mysql-5.0.76
         =virtual/mysql-5.1 )
    >=kde-base/kdelibs-4.3[opengl?,
                   semantic-desktop?]
    sys-libs/zlib
    x11-libs/qt-script
    >=x11-libs/qtscriptgenerator-0.1.0
\end{BVerbatim}
\caption{Fragment of the runtime dependencies of the Amarok music player.}\label{fig:amarokruntime}
\end{wrapfigure}

@@

@@

The dataset collected is available at
\begin{align}
\mbox{\href{http://datahub.io/dataset/gentoo-dependency-graph}{\texttt{http://datahub.io/dataset/gentoo-dependency-graph}}}
\end{align}
as compressed GML graphs for every snapshot. Raw data is available at
\begin{align}
\mbox{\url{http://www.utwente.nl/mb/iebis/staff/amrit/timeline.squashfs}}
\end{align}
as a SquashFS compressed filesystem. The SquashFS format allows one to mount the compressed data as a read-only disk drive and operate directly on the dataset. It contains the pre-processed ebuild files and the file \verb|graphs.bip| that contains all of the graphs in a custom format specialized for quick processing of large graphs over time. A C++ toolkit to work with the \verb|graphs.bip| file is available at \url{https://github.com/Recmo/depgraph}.

@@

From the entire list of packages, a few well-known (at least according to the author) packages where selected. The selection criteria was that the package must not have existed before (around) 2004, because the Gentoo Portage database was still too immature then, and the package must have gained a considerable number of dependers since its introduction. In table \ref{tbl:results}, we list some the packages along with their parameters. In these cases the adjusted coefficient of determination $\overline{R}^2$ was more than $99\%$.

Special tools where developed to extract the time series of the number of adopters $A_t$ for a given package. This time series was then fit to \eqref{eq:bass} using Mathematica's \texttt{NonLinearModelFit}. The goodness-of-fit was analysed using an ANOVA table and calculated using the adjust coefficient of determination $\overline{R}^2$. The parameters were extracted from the fit, and confidence intervals were calculated under assumptions of normality. In table \ref{tbl:results} the relevant parameters are presented with their mean value and a $95\%$ confidence interval. Since normality was assumed, the confidence intervals ignore the $p \ge 0$ and $q \ge 0$ constraints.

The plots were drawn using a red line for the model and shades of red for the confidence bands. The red line was drawn according to \eqref{eq:bass}, with the mean values used as parameters. Then single value predictions confidence bands were calculated for $90\%$, $95\%$, $99\%$ and $99.9\%$ confidence and were drawn in progressively darker shades of pink. They represent a prediction for where a single additional value would likely fall. According to the model and the uncertainty introduced by the fit, there is a $90\%$ chance that it will fall in the innermost band, $95\%$ chance that it would falil in the second band, etc. If the data fits the model properly, one would expect to see $90\%$ of the points in the inner band. Finally, the empirical data points were plotted as black dots, connected with a thin vertical line - the residual error of the model.

It turns out the packages can be classified as mostly imitator driven, mostly innovator driven and an exceptional case of growth and demise. They will be discussed in order.

@@

The first package we consider is \verb|git|, a modern revision control system that shows an imitator driven adoption. It's growth can be seen in figure \ref{fig:git}, the corresponding statistics are in table \ref{tbl:results}. The package first appeared just before 2005, it had around ten packages depending on it in 2006, twenty in 2008 and is currently used by almost three hundred packages. According to the Bass model, it will continue to grow to approximately 750 users. The innovator inflow is only $0.2\%$ of the potential market per year, so one would expect $0.002\cdot (750 - 300) = 1$ user to adopt \verb|git| out of sheer innovation. Taking the analogy with persons, if someone from the $450$ current non-users where to meet a random person from the entire $750$ market, there is a $\frac{300}{750}= 40\%$ chance of meeting a user which can convince him/her to start using \verb|git|. The chance of this happening is the imitator inflow $q = 0.73$. Therefore, the total number of users \verb|git| can expect to gain from imitation this year is $450\cdot 40\%\cdot 0.73 = 131$. Very much imitator driven!

@@

A typical example of innovator driven growth is given by \verb|libmad|. The model is fitted resulting in figure \ref{fig:libmad}. Again, the data is neatly explained by a Bass diffusion process, in particular the rapid steep growth and the stable user base afterwards. The name is an acronym for ``library for MPEG Audio Decoding'' and the package provides a high quality mp3 decoder for use in multimedia applications. This might also explain the rapid growth of its adoption: multimedia applications can benefit a lot from good quality mp3 support.

@@

\subsection{Hypothesis Testing}

The above examples of Bass diffusion growth suggests the central hypotheses of this article: \emph{software dependencies are a Bass innovation diffusion process.} This hypotheses is tested using a methodic approach: a random selection of ten software packages is taken and their dependee growth is fitted to the Bass diffusion model.

The Gentoo portage database contains over twenty thousand packages, but only a fraction of these packages are suitable. About six thousand packages where introduced before 2004, where the growth of Gentoo itself may be confounded. Of the remaining packages, the vast majority do not have a number of other packages depedending on it, these would for example be applications (as opposed to libraries) or unpopular/niche libraries. These are filtered out by setting a lower bound for the maximum number of dependees a package has during its lifetime. Then there are infrastructural packages in Gentoo, that have no corresponding open source projects, but are used manage the pacakge database. In summary:

\begin{enumerate}
	\item Start with all the packages in the Gentoo portage database. (21\,380 packages)
	\item Filter out the packages that existed before 2004-01-01. (14\,706 remaining)
	\item Filter out packages that do not have at least 20 dependees at some point. (324 remaining)
	\item Filter out packages labeled ``-virtual'', ``-virtuals'', ``-proto'' or ``-meta''. (259 remaining)
\end{enumerate}

After this initial automated filtering, packages are randomly selected and manually tested against further selection criteria. If a package is rejected a new one is randomly selected and the process is repeated untill ten packages have been found. The final criteria are:

\begin{enumerate}
	\item[A.] The dependeee growth does not contain strong discontinuities. It is very uncommon that many independent software project suddenly, within the time resolution of one month, decide to start or stop using a particular library. It is however common that maintainers of Gentoo Portage do reorganizations or mass additions/removals of packages that product such discontinuities. Since they do not reflect actual adoption they are noise and are filtered out.
	\item[B.] The package corresponds to an open source software project. Some of the packages represent closed source software projects or non-software projects. These are not part of the scope and are therefore excluded.
	\item[C.] The package has not been forked of or factored out of another package. Some libraries where initially part of a different library that got refactored and split up in separate libraries and packages. Since projects may have been depedending on the initial library and simply switch over to the refactored one, this can not be considered adoption of the library. Such packages are therefore excluded.
	\item[D.] The first included version should be close to a '1.0' release. The Bass diffusion hypotheses is about the entire growth of an open source software project, it is therefore necessary to have dependee growth data from close to the beginning of the project. Package that had an initial public release long before inclusion in Gentoo Portage are therefore excluded.
\end{enumerate}

These criteria are verified by consulting the information available through url \eqref{url:pkg}, in particular: the changelog of the ebuild file, any readme or changelog included in the earliest available version of the source code, the projects revision history and the projects homepage.

The manutally rejected packages are listed in table \ref{tbl:rejected}.

C: Xorg, KDE and Qt. Interesting. They actually show Bass diffusion. Refactor/version upgrade has some diffusion to it?

%NOTE: Some rejected packages would still be good fits for the Bass diffusion model.

@@

@@

\begin{table}
\small\centering
\caption{Results of fitting the Bass model.}\label{tbl:testresults}
\begin{tabular}{lr@{ $\pm$}rr@{ $\pm$}rr@{ $\pm$}rr}
\toprule
Package & \multicolumn{2}{c}{$p$} & \multicolumn{2}{c}{$q$} & \multicolumn{2}{c}{$M$} & $\overline{R}^2$\\
\midrule
% 34, 65, 218, 77, 157, 240, 242, 17, 206, 143
\texttt{dev-dotnet/gconf-sharp}   & 0.168 & 0.212 & 0.754 & 0.434 &  21 &   1 & 99.61\%\\
\texttt{dev-haskell/cabal}        & 0.055 & 0.074 & 0.154 & 0.380 & 246 & 438 & 98.55\%\\
\texttt{dev-libs/libunique}       & 0.047 & 0.099 & 0.136 & 0.402 & 287 & 732 & 98.58\%\\
\texttt{dev-python/lxml}          & 0.007 & 0.009 & 0.488 & 0.217 &  53 &  53 & 98.82\%\\
\texttt{dev-python/simplejson}    & 0.011 & 0.029 & 0.463 & 0.459 &  97 & 148 & 98.98\%\\
\texttt{dev-vcs/git}              & 0.000 & 0.006 & 0.841 & 0.144 & 734 & 369 & 99.24\%\\
\texttt{dev-vcs/mercurial}        & 0.001 & 0.005 & 0.517 & 0.259 & 174 & 481 & 98.24\%\\
\texttt{gnome-base/gnome-keyring} & 0.026 & 0.036 & 0.833 & 0.133 &  42 &   2 & 99.66\%\\
\texttt{media-libs/libcanberra}   & 0.055 & 0.117 & 1.436 & 0.439 &  37 &   3 & 99.68\%\\
\texttt{x11-misc/xdg-utils}       & 0.050 & 0.043 & 0.257 & 0.242 & 115 & 127 & 99.56\%\\
\bottomrule
\end{tabular}
\end{table}

Five packages are good fits with an $\overline{R}^2 > 99\%$. Of the remaining five, \texttt{dev-dotnet/lxml} and \texttt{dev-cvs/mercurial} can be explained by discretization noise. The number of dependees is integer and can therefore only rise or fall in descrete steps, this introduces additional variance. It also appears to happen in \texttt{dev-haskell/cabal}, but this is an illusion since the steps in the growth are in the order of tens of packages. The reasons for these steps are unknown and likely related to Gentoo internal matters. \texttt{dev-python/simplejson} is barely under the mark, and appears to have two distinct growth spurts. It is unknonw wether these growth spurts are an artifact of the Gentoo Portage database, or if they represent two specific growth phases in the project. The first spurt coincides with the \texttt{2.0} release, but no such correlation can be found for the second spurt.

\texttt{dev-libs/libunique} is the only one case that clearly does not fit the Bass diffusion model, and it appears to be for the same reasons that \texttt{xulrunner} does not fit.
